---
title: "Functional programming in R"
date: 2018-10-07
author: "Lucas A. Nell"
toc: true
filters:
  - webr
---

This document is based on a presentation I did for the grad student
organization for the Department of Integrative Biology, UW--Madison in Fall
2018. I've made a few changes to hopefully make it more transparent as a
stand-alone document.

A few of the code blocks below (marked with a **Run Code** button) are live —
edit them and rerun to see what happens, no R installation required.

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Why use functions?

Two main advantages over copy and paste:

1. Create fewer errors
2. Improve readability of code

## Consider the following example

Small errors are easy to make and can be annoying to find.

```{r show_errors, error=TRUE}
lm_mpg <- lm(mpg ~ factor(cyll), mtcars)
```
```{r show_errors2, error=TRUE}
lm_hp <- lm(hp ~ factor(cyl), mrcars)
```
```{r show_errors3, error=TRUE}
lm_disp <- lm(mpg ~ factor(cyl), mtcars)
```

The problem is even worse when you have lots of copying.

```{r show_many_models, eval=FALSE}
lm(mpg ~ cyl + disp + hp + drat, mtcars)
lm(mpg ~ cyl + disp + hp + wt, mtcars)
lm(mpg ~ cyl + disp + drat + wt, mtcars)
lm(mpg ~ cyl + hp + drat + wt, mtcars)
lm(mpg ~ disp + hp + drat + wt, mtcars)
lm(disp ~ mpg + cyl + hp + drat, mtcars)
lm(disp ~ mpg + cyl + hp + wt, mtcars)
lm(disp ~ mpg + cyl + drat + wt, mtcars)
lm(disp ~ mpg + hp + drat + wt, mtcars)
lm(disp ~ cyl + hp + drat + wt, mtcars)
lm(hp ~ mpg + cyl + disp + drat, mtcars)
lm(hp ~ mpg + cyl + disp + wt, mtcars)
lm(hp ~ mpg + cyl + drat + wt, mtcars)
lm(hp ~ mpg + disp + drat + wt, mtcars)
lm(hp ~ cyl + disp + drat + wt, mtcars)
```

```{r fixed_errors_fxn, echo=FALSE}
cyl_model <- function(y) {
  form <- as.formula(paste(y, "~ factor(cyl)"))
  mod <- lm(form, mtcars)
  mod$call$formula <- as.formula(mod$terms)
  return(mod)
}
```

### Which is better?

```{r non_fixed_errors, eval = FALSE}
lm_mpg <- lm(mpg ~ factor(cyl), mtcars)
lm_disp <- lm(disp ~ factor(cyl), mtcars)
lm_hp <- lm(hp ~ factor(cyl), mtcars)
lm_drat <- lm(drat ~ factor(cyl), mtcars)
lm_wt <- lm(wt ~ factor(cyl), mtcars)
lm_qsec <- lm(qsec ~ factor(cyl), mtcars)
lm_vs <- lm(vs ~ factor(cyl), mtcars)
lm_am <- lm(am ~ factor(cyl), mtcars)
lm_gear <- lm(gear ~ factor(cyl), mtcars)
lm_carb <- lm(carb ~ factor(cyl), mtcars)
```

or

```{r fixed_errors}
y_pars <- c("mpg", "disp", "hp", "drat",
            "wt", "qsec", "vs", "am",
            "gear", "carb")
all_lm <- lapply(y_pars, cyl_model)
```

## Some R basics

### Basics of functions in R

```{webr-r}
#| label: fxn_intro
#| autorun: true
subtract <- function(x, y = 1) {
  z <- x - y
  return(z)
}
subtract(1:3, 4:6)
subtract(1:3)
subtract(y = 1:3, x = 4:6)
```

### Flexibility of lists

```{webr-r}
#| label: lists_are_great
#| autorun: true
x <- numeric(2)
x[[1]] <- matrix(0, 0, 0)
```

```{webr-r}
#| label: lists_are_great2
#| autorun: true
x <- as.list(numeric(3))
x[[1]] <- matrix(0, 0, 0)
x[[2]] <- data.frame()
x[[3]] <- runif(3)
x
```

### The `apply` functions

- Allows you to apply a function to multiple inputs.
- `lapply` outputs a list, `sapply` coerces to an array.

```{webr-r}
#| label: lapply_ex
#| autorun: true
lapply(4:5, function(i) 1 + i)
sapply(4:5, function(i) 1 + i)
```

### For loops

- Especially useful when one iteration's result depends on the previous iteration.
- Changes existing object(s).

```{webr-r}
#| label: for_loops2
#| fig-width: 8
#| fig-height: 4
#| autorun: true
set.seed(3)
par(mar = c(4, 4, 1, 1))
x <- numeric(100)
x[1] <- 10
for (t in 2:length(x)) {
  x[t] <- x[t-1] + rnorm(1)
}
plot(x, type = "l", lwd = 2)
```

## General process to "functionalize" code

1. Break problem into smaller sub-problems.
2. For each sub-problem, write a function.
3. For writing each function...
  1. The main function code will include the commonalities between all situations.
  2. Features that aren't common should be input to the function as arguments.

## Example #1: Cleaning weird files

Suppose we have a folder full of CSV files like this:

```{r weirddata_data, echo = FALSE}
files <- c("## Data provided by X

Ozone,Solar.R,Wind,Temp,Month,Day
41,190,7.4,67,5,1
NA,NA,14.3,56,5,5
--- instrument error
28,NA,14.9,66,5,6
23,299,8.6,65,5,7
--- instrument error
NA,194,8.6,69,5,10

## Year observed: 1990",
"## Data provided by Y

Ozone,Solar.R,Wind,Temp,Month,Day
7,NA,6.9,74,5,11
--- instrument error
11,290,9.2,66,5,13
14,274,10.9,68,5,14
18,65,13.2,58,5,15
--- instrument error
--- instrument error
6,78,18.4,57,5,18
30,322,11.5,68,5,19
11,44,9.7,62,5,20

## Year observed: 1991
")
cat(files[1])
```

#### Problems:

1. Remove unnecessary lines from each file.
2. Create a single data frame from multiple cleaned files.

#### Input information:

1. Vector of file names (`file_names`)

```{r weirddata_info, eval=FALSE}
file_names <- c("file1.csv", "file2.csv")
```

### Clean a single CSV file to a string

```{r weirddata3, eval=FALSE}
clean_str <- function(file_name) {
  lines <- readLines(file_name)
  lines <- lines[!grepl("^\\#\\#|^--", lines) &
                   lines != ""]
  cleaned_str <- paste(lines, collapse = "\n")
  return(cleaned_str)
}
```

### Clean multiple files then combine them into a single data frame:

```{r weirddata4, eval=FALSE}
clean_df <- function(file_names) {
  cleaned_strs <- lapply(file_names, clean_str)
  data_frames <- lapply(cleaned_strs, function(s) read.csv(text = s))
  combined_df <- do.call(rbind, data_frames)
  return(combined_df)
}
head(clean_df(file_names))
```
```{r weirddata5, eval=TRUE, echo=FALSE}
df_list <- lapply(strsplit(files, "\n"),
       function(lines) {
         lines <- lines[!grepl("^\\#\\#|^--", lines) &
                 lines != ""]
         cleaned_str <- paste(lines, collapse = "\n")
         read.csv(text = cleaned_str)
       })
head(do.call(rbind, df_list))
```

## Example #2: Fitting lots of models

How can we simplify this?

```{r lotsa_models_bad, eval = FALSE}
lm(mpg ~ cyl + disp + hp + drat, mtcars)
lm(mpg ~ cyl + disp + hp + wt, mtcars)
lm(mpg ~ cyl + disp + drat + wt, mtcars)
lm(mpg ~ cyl + hp + drat + wt, mtcars)
lm(mpg ~ disp + hp + drat + wt, mtcars)
lm(disp ~ mpg + cyl + hp + drat, mtcars)
lm(disp ~ mpg + cyl + hp + wt, mtcars)
lm(disp ~ mpg + cyl + drat + wt, mtcars)
lm(disp ~ mpg + hp + drat + wt, mtcars)
lm(disp ~ cyl + hp + drat + wt, mtcars)
lm(hp ~ mpg + cyl + disp + drat, mtcars)
lm(hp ~ mpg + cyl + disp + wt, mtcars)
lm(hp ~ mpg + cyl + drat + wt, mtcars)
lm(hp ~ mpg + disp + drat + wt, mtcars)
lm(hp ~ cyl + disp + drat + wt, mtcars)
```

#### Problems:

1. Create all necessary formulas for each of the multiple Ys.
2. Fit `lm` based on each of the created formulas.

#### Input information:

1. Vector of Y variables (`Ys`)
2. Vector of possible X variables (`Xs`)
3. Number of X variables to include in each model (`n_Xs`)

```{webr-r}
#| label: lotsa_models_info
#| autorun: true
Ys <- c("mpg", "disp", "hp")
Xs <- c("mpg", "cyl", "disp", "hp", "drat", "wt")
n_Xs <- 4
```

### Make vector of all necessary formulas:

```{webr-r}
#| label: lotsa_models
#| autorun: true
make_forms <- function(y, Xs, n_Xs) {
  poss_Xs <- Xs[Xs != y]
  n_poss_Xs <- length(poss_Xs)
  # All possible combinations:
  combs <- combn(n_poss_Xs, n_Xs, simplify = FALSE)
  # Change to names:
  names_ <- lapply(combs, function(x) poss_Xs[x])
  # Combine each set to single RHS of formula:
  rhs <- sapply(names_, paste, collapse = " + ")
  # Whole formulas as strings:
  form_strings <- paste(y, "~", rhs)
  # Convert to formulas:
  forms <- sapply(form_strings, as.formula,
                  USE.NAMES = FALSE)
  return(forms)
}
```

### Fit `lm()` based on a single formula:

```{webr-r}
#| label: lotsa_models2
#| autorun: true
single_mod <- function(form) {
  # Fit lm:
  mod <- lm(form, mtcars)
  # Make model print prettier:
  mod$call$formula <- as.formula(mod$terms)
  return(mod)
}
```

### Put both steps together:

```{webr-r}
#| label: lotsa_models3
#| autorun: true
fit_models <- function(Ys, Xs, n_Xs) {
  forms <- lapply(Ys, make_forms,
                  Xs = Xs, n_Xs = n_Xs)
  forms <- c(forms, recursive = TRUE)
  lms <- lapply(forms, single_mod)
  return(lms)
}
model_fits <- fit_models(Ys, Xs, n_Xs)
model_fits[[1]]
```

## More information

- T Mailund (2017). *Functional Programming in R*. [doi: 10.1007/978-1-4842-2746-6](
  https://link.springer.com/book/10.1007%2F978-1-4842-2746-6)
  - Free through UW
- [Functional Programming in R with purrr (towardsdatascience.com)](
https://towardsdatascience.com/functional-programming-in-r-with-purrr-469e597d0229)
- [Functional Programming (in Advanced R)](
  http://adv-r.had.co.nz/Functional-programming.html)
